The dependence of spiral disk morphology on the star formation-stellar mass relation

### 2014-09-16 paper/mainsequence_morph.tex
Incorporated comments by RS, OIW, TM

\begin{abstract}
We measure the mass-star formation relation in disk galaxies at $z<0.05$, using Galaxy Zoo morphologies to examine different populations of spirals as classified by their kiloparsec-scale structure. We examine the number of spiral arms, their average pitch angle, and the presence of a galactic bar in the disk, and show that both the slope and dispersion of the $M_\star$-SFR relation is constant when varying all the above parameters. We also show that mergers (both major and minor), which represent the strongest conditions for increases in star formation at a constant mass, only drive the galaxies off the main relation by $\sim0.3$~dex; this is a significant reduction over the increase seen in merging systems at $z>1$. Of the galaxies lying significantly above the $M_\star$-SFR relation in the local Universe, more than $50\%$ are mergers. We interpret this as evidence that the spiral arms, which are imperfect reflections of the galaxy's current gravitational potential, are either fully independent of the basic quenching relation or are completely overwhelmed by the combination of outflows and feedback. The arrangement of the star formation can be changed (as demonstrated by the filling factor of the disk), but the system \emph{as a whole} regulates itself even in the presence of strong dynamical forcing. 
\end{abstract}

@@

\begin{figure*}
\includegraphics[angle=0,width=7.0in]{figures/ms_arms_number.pdf}
\caption{Total star formation rate as a function of stellar mass; grayscale colors are the distribution of all star-forming galaxies in SDSS from the MPA-JHU DR7 catalog. Colored points in each panel show spiral galaxies with 1, 2, 3, 4, more than four, or ``uncertain'' numbers of spiral arms per galaxy, as identified by GZ2 morphology flags. Dotted lines show the weighted least-squares linear fit to each population as split by arm multiplicity; the solid line is the fit to all of the star-forming galaxies. 
\label{fig-number}}
\end{figure*}

Figure~\ref{fig-number} overplots SFR as a function of $M_\star$ for disk galaxies separated by their arm multiplicity. The GZ2 data separates disk galaxies with visible spiral arms into categories of 1, 2, 3, 4, or more than four spiral arms; there is also an additional option if the number of spiral arms cannot be accurately determined. There are a total of 14,179 galaxies with flags for arm multiplicity (indicating a high level of agreement among classifiers) in the GZ2 data. Of those, two-armed spirals are by far the most common, comprising 85\% of the total. This is also the only multiplicity for which significant numbers of galaxies at $M_\star<10^9~M_\sun$ are detected.

The fit to the star-forming main sequence for two-armed spirals tighly follows that of that for all spiral galaxies, with both the slope and offset of the linear fits completely consistent with each other (Table~\ref{tbl-fits}). Similarly, the fits for 3, 4, and $5+$-armed spirals are also consistent with the main relation (although all three have much lower numbers which affect the quality of the fit).

The one-armed spirals present an interesting case, with the majority of the population lying slightly above the fit to all star-forming spirals. This is consistent with the work of \citet{cas13}, who showed that one-armed spirals in GZ2 are robust indicators of close interactions at projected distances of $r_p < 50~h^{-1}$~kpc. The underlying reason is that many ``one-armed spirals'' are in fact caused by bridges or tidal tails from interactions with a nearby companion; we discuss the likely role of merging/interacting galaxies in Figure~\ref{fig-mergers}.

Finally, galaxies with an uncertain number of spiral arms (although also with relatively low numbers) fall slightly below the mean star-forming main sequence relation. One possibility is that this is a consequence of galaxies that have already begun the process of quenching. As a result, the star-formation rate is depleted and the contrast of spiral arms against the rest of the disk (which for visual identification is at least in part dependent on bright knots of recent star formation) makes the multiplicity more difficult to identify. However, the $(u-r)$ colors of these galaxies (uncorrected for dust extinction) are on average significantly bluer than green-valley late-types as identified in \citet{sch14}.

@@

It should also be noted that the galaxies in GZ2 flagged as a function of pitch angle are not representative of the true vote distribution. The raw numbers in Figure~\ref{fig-winding} suggest that most spiral galaxies are either tightly or loosely wound. In fact, the plurality classification for most galaxies is for medium-winding; the spread in votes is typically large, though, and so users rarely agree on the medium option at the 80\% level which sets the flag. Plotting the same star-forming main sequence diagram as a function of $p_{\rm medium}$, which more accurately tracks the pitch angle distribution, also closely follows the main star-forming main sequence. % Figure ms_arms_winding_weighted_pmed.pdf shows this.

@@

@@

The agreement of all sub-varieties of star-forming galaxies is supported by the close agreement to the linear fits to the data for all well-sampled categories (Table~\ref{tbl-fits}). This tracks only the slope and offset of the distribution, however, and not its width. We thus also compared the sample standard deviation ($\sigma_{SFR}$) as a function of mass of the star-forming galaxy population to its various subsamples. The value of $\sigma_{SFR}$ monotonically decreases with mass over the range $6.0<\log(M/M_\odot)<11.5$. For all populations examined in this paper, though, the widths of their distributions are consistent with the broader population. % Figure to be shown in Appendix?

@@

>>>>

\begin{abstract}
We measure the mass-star formation relation in disk galaxies at $z<0.05$, using Galaxy Zoo morphologies to examine different populations of spirals as classified by their kiloparsec-scale structure. We examine the number of spiral arms, their average pitch angle, and the presence of a galactic bar in the disk, and show that both the slope and dispersion of the $M_\star$-SFR relation is constant when varying all the above parameters. We also show that mergers (both major and minor), which represent the strongest conditions for increases in star formation at a constant mass, only boost the SFR above the main relation by $\sim0.3$~dex; this is a significant reduction over the increase seen in merging systems at $z>1$. Of the galaxies lying significantly above the $M_\star$-SFR relation in the local Universe, more than $50\%$ are mergers. We interpret this as evidence that the spiral arms, which are imperfect reflections of the galaxy's current gravitational potential, are either fully independent of the basic quenching relation or are completely overwhelmed by the combination of outflows and feedback. The arrangement of the star formation can be changed (as demonstrated by the filling factor of the disk), but the system \emph{as a whole} regulates itself even in the presence of strong dynamical forcing. 
\end{abstract}

@@

\begin{figure*}
\includegraphics[angle=0,width=7.0in]{figures/ms_arms_number_weighted_contour.pdf}
\caption{Total star formation rate as a function of stellar mass; grayscale colors are the distribution of all star-forming galaxies in SDSS from the MPA-JHU DR7 catalog. Coloured contours in each panel show spiral galaxies weighted by the GZ2 likelihoods of hosting 1, 2, 3, 4, more than four, or ``uncertain'' numbers of spiral arms, respectively. Dotted lines show the weighted least-squares linear fit to each population as split by arm multiplicity; the solid line is the fit to all of the star-forming galaxies. 
\label{fig-number}}
\end{figure*}

Figure~\ref{fig-number} overplots SFR as a function of $M_\star$ for disk galaxies separated by their arm multiplicity. The GZ2 data separates disk galaxies with visible spiral arms into categories of 1, 2, 3, 4, or more than four spiral arms; there is also an additional option if the number of spiral arms cannot be accurately determined. For this analysis, galaxies in each mass/SFR bin is weighted by the vote fraction for the morphology being tested. Two-armed spirals are the most common multiplicity in the GZ2 sample.

The fit to the star-forming main sequence for the six sub-samples of spiral galaxies tightly follows that of that for the total disk population. Both the slopes and offsets of each linear fit are completely consistent within the formal fitting errors (Table~\ref{tbl-fits}). There are too few galaxies with reliable spiral morphologies at lower masses to do a reliable fit; however, the same relation is reproduced for all multiplicities at $10^8<M_\star/M_\sun<10^{11}$.

The one-armed spirals present an interesting case, with the fit to the weighted population lying slightly \emph{above} that of all star-forming spirals. This is consistent with the work of \citet{cas13}, who showed that one-armed spirals in GZ2 are robust indicators of close interactions at projected distances of $r_p < 50~h^{-1}$~kpc. The underlying reason is that many ``one-armed spirals'' are in fact caused by bridges or tidal tails from interactions with a nearby companion; we discuss the likely role of merging/interacting galaxies in Figure~\ref{fig-mergers}.

The two-armed spirals and those with uncertain multiplicities are the only morphologies that extend slightly below the SFMS. One possibility is that this is a consequence of galaxies that have already begun the process of quenching. As a result, the star-formation rate is depleted and the contrast of spiral arms against the rest of the disk (which for visual identification is at least in part dependent on bright knots of recent star formation) makes the multiplicity more difficult to identify. However, the $(u-r)$ colors of these galaxies (uncorrected for dust extinction) are on average significantly bluer than green-valley late-types as identified in \citet{sch14}.

We have repeated the analysis above for the sub-sample of disk galaxies for which the spiral multiplicity is determined with high confidence ($p_\textrm{arms~number} > 0.8$) by GZ2, thus eliminating ``intermediate'' galaxies for which the morphology is uncertain. These galaxies ($N=14,148$) is dominated by two-armed spirals, which are the only multiplicity for which significant numbers of galaxies at $M_\star<10^9~M_\sun$ are detected. Our results for the SFMS are qualitatively the same as when using the weighting scheme, although we note that there are too few examples ($N<10$) of either three- or four-armed spirals for a reliable fit. The offset of the one-armed spirals above the SFMS is also significantly more pronounced when using high-confidence morphologies.

@@

It should also be noted that the galaxies in GZ2 flagged as a function of pitch angle are not representative of the true vote distribution. The raw numbers in Figure~\ref{fig-winding} suggest that most spiral galaxies are either tightly or loosely wound. In fact, the plurality classification for most galaxies is for medium-winding; the spread in votes is typically large, though, and so users rarely agree on the medium option at the 80\% level which sets the flag. We have extended this analysis by plotting the same star-forming main sequence diagram as a function of $p_\textrm{arms~winding}$, which more accurately tracks the pitch angle distribution. In this case, it also closely follows the main star-forming main sequence (Figure~\ref{fig-winding_weighted}).

@@

The agreement of all sub-varieties of star-forming galaxies is supported by the close agreement to the linear fits to the data for all well-sampled categories (Table~\ref{tbl-fits}). This tracks only the slope and offset of the distribution, however, and not its width. We thus also compared the sample standard deviation ($\sigma_{SFR}$) as a function of mass of the star-forming galaxy population to its various subsamples. The value of $\sigma_{SFR}$ monotonically decreases with mass over the range $6.0<\log(M/M_\odot)<11.5$. For all populations examined in this paper, though, the widths of their distributions are consistent with the broader population (Figure~\ref{fig-sigma}).

\begin{figure}
%\includegraphics[angle=0,width=7.0in]{figures/ms_bar.pdf}
\includegraphics[angle=0,width=3.5in]{figures/sigma_mstar.pdf}
\caption{Width of the star-forming main sequence ($\sigma$) as a function of stellar mass, as measured by the sample standard deviation. Black points represent the entire star-forming population. The various disk subsamples are overplotted as lines; colours are the same as the respective plots in Figures~\ref{fig-number}-\ref{fig-bar} and \ref{fig-mergers}. Morphological categories or mass ranges with fewer than 10 galaxies/bin are not plotted; this includes all mass ranges for 3-, 4-, and more than 4-armed spirals.
\label{fig-sigma}}
\end{figure}

@@

@@

\appendix
\begin{figure*}
\includegraphics[angle=0,width=7.0in]{figures/ms_arms_winding_weighted.pdf}
\caption{Total star formation rate as a function of stellar mass; grayscale colors are the same as in Figure~\ref{fig-winding}. From left to right: colormaps show spiral galaxies weighted by the GZ2 vote fractions for ``tight'', ``medium'', and ``loose'' winding spiral arms. Dotted lines show the weighted least-squares linear fit as split by pitch angle; the solid line is the fit to all star-forming galaxies. 
\label{fig-winding_weighted}}
\end{figure*}
